---
title: "Multilevel linear Models"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---


```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, prompt = TRUE, warning = FALSE, message = FALSE, comment=NA)
library(lme4)
library(ggplot2)
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(nlme)
library(brms)

schools <- c('1308', '1946', '7276', '9198', '2917', '9104', '4042', '7635', '8188', '6464', '5838', '9158', '5815', '6808', '6578', '4350')
Df <- MathAchieve %>% 
  select(school = School,
         minority = Minority,
         sex = Sex,
         ses = SES,
         mathach = MathAch) %>% 
  filter(school %in% schools) %>% 
  mutate(school = as.character(school))

set.seed(10101)
mu <- rnorm(16, mean=500, sd=50)
f <- function(mu) rnorm(250, mean=mu, sd=25)
fake_rt <- do.call(cbind, lapply(mu, f)) %>% 
  as.data.frame() %>% 
  as_tibble() %>% 
  rename_all(funs(str_replace(., "V", "subject"))) %>% 
  gather(subject, reaction_time, subject1:subject16) 

#write_csv(fake_rt, '../data/fake_rt.csv')

```

# Multilevel data: Example 1

Reaction time as a function of sleep deprivation.
```{r}
ggplot(sleepstudy,
       aes(x=Days, y=Reaction, col=Subject)
) + geom_point() +
  stat_smooth(method='lm', se=F) +
  facet_wrap(~Subject) +
  guides(col=F)
```

# Multilevel data: Example 2

Mathematical achievement as function of socio-economic status.
```{r}
ggplot(Df,
       aes(x=ses, y=mathach, col=school)
) + geom_point() +
  stat_smooth(method='lm', se=F) +
  facet_wrap(~school) +
  guides(col=F)
```



# Example: Reaction time and math achievement

-   In this problem, we have $J$ subject. For subject $j$, we have
    $n_j$ data points.

-   In observation $i$ from subject $j$, their number of days without sleep is $x_{ji}$ and the reaction time is $y_{ji}$.

-   A multilevel model for this data is $$\begin{aligned}
    y_{ji} &\sim N(\alpha_{j} + \beta_j x_{ji},\sigma^2),\\
    \alpha_j &\sim N(a,\tau_a^2),\\
    \beta_j &\sim N(b,\tau_b^2).\end{aligned}$$

# Example: Reaction time and math achievement

-   The model $$\begin{aligned}
    y_{ji} &\sim N(\alpha_{j} + \beta_j x_{ji},\sigma^2),\\
    \alpha_j &\sim N(a,\tau_a^2),\\
    \beta_j &\sim N(b,\tau_b^2),\end{aligned}$$ can be re-written
    $$y_{ji} = \underbrace{(a + \eta_j)}_{\text{ $\alpha_j$}} + \underbrace{(b + \zeta_j)}_{\text{ $\beta_j$}} x_{ji} + \epsilon_{ji},$$
    or
    $$y_{ji} = \underbrace{a + bx_{ji}}_{\text{Fixed effect}} + \underbrace{\eta_j + \zeta_j x_{ji}}_{\text{Random effect}} + \epsilon_{ji},$$
    where
    $$\eta_j \sim N(0,\tau_a^2),~ \zeta_j \sim N(0,\tau_b^2),~ \epsilon_j \sim N(0,\sigma^2).$$

    



# Example:  Reaction time and math achievement

-   In the model just described, $a$ and $b$ are the general regression
    coefficients.

-   The variance $\tau_a^2$ tells us how much variation in the intercept
    term there is across schools. The variance $\tau_b^2$ tells us how
    much variation in the slope term there is across schools.

-   For example, 95% and 99% of the intercepts for individual schools
    will be in the ranges
    $$a \pm 1.96 \times \tau_a,\quad  a \pm 2.56 \times \tau_a,$$
    respectively. Likewise, 95% and 99% of the slope terms for schools
    will be in the ranges
    $$b \pm 1.96 \times \tau_b,\quad b  \pm 2.56 \times \tau_b.$$

